\documentclass{article}
\usepackage[left=3cm,right=3cm,top=3cm,bottom=3cm]{geometry}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[OT1]{fontenc}
\usepackage{hyperref}
\usepackage[inference]{semantic}
\usepackage{graphicx}
\usepackage{amsfonts,amsmath}
\usepackage{cleveref}
\usepackage{scalerel}
\usepackage{multicol}
\usepackage{xspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{natbib}
\usepackage{booktabs}

\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{colonequals}
\usepackage{xspace}
% Commands for correct abbreviations


\usepackage{graphicx}
\usepackage{bussproofs}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,chains}

\EnableBpAbbreviations

\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows.meta}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}

\usepackage{dsfont}

\renewcommand{\to}{\Rightarrow}
\newcommand{\TT}{\mathcal{T}}
\newcommand{\MM}{\mathcal{M}}
\newcommand{\KK}{\mathcal{K}}
\newcommand{\UU}{\mathcal U}

\newtheorem{exercise}{Exercise}

\begin{document}

\title{Natural Deduction}
\author{Dominik Kirst}

\date{LMFI Proof Theory Course\\October 17th, 2025}

\maketitle

\begin{exercise}[ND Derivations]
	Give derivations for $\vdash \neg (A\Leftrightarrow \neg A))$ and $\vdash_c \neg (\forall x \neg A)\to \exists x A$.
\end{exercise}

\begin{exercise}[Weakening]
	Recall the weakening property that $\Delta\vdash A$ whenever $\Gamma\vdash A$ and $\Gamma\subseteq \Delta$.
	\begin{enumerate}
		\item
		Using induction on $\Delta\vdash A$, give the inductive hypothesis for each rule.
		\item
		Establish the weakening property for all propositional rules of ND.
		\item
		What do you observe in the case of the quantifier rules?
		\item
		Establish the renaming property that $\Gamma[y/x]\vdash A[y/x]$ whenever $\Gamma\vdash A$ and $y$ is completely fresh.
		\item
		Use the renaming property to finish the weakening proof for the quantifier rules.
	\end{enumerate}
\end{exercise}

\begin{exercise}[Equivalence of LK and NK]
	Show that if $\Gamma\vdash \Delta$, then $\Gamma\vdash_c\bigvee\Delta$.
\end{exercise}

\begin{exercise}[Double-negation Translation]
	For the translation $A^*$ as defined in the lecture show:
	\begin{enumerate}
		\item
		Show that $\vdash_c A\Leftrightarrow A*$ by induction on $A$.
		\item
		Show that $\vdash \neg \neg A^*\to A^*$ by induction on $A$.
		\item
		Show that $\Gamma\vdash_c A$ implies $\Gamma^*\vdash A^*$.
		\item
		Can you even show that $\Gamma\vdash_c A$ implies $\Gamma^*\vdash_m A^*$?
	\end{enumerate}
\end{exercise}

\begin{exercise}[Henkin Completeness]
	Extend the Henkin completeness proof to the full syntax:
	\begin{enumerate}
		\item
		Add respective clauses to the Lindenbaum lemma
		\item 
		Prove the extended Lindenbaum lemma. Hint: you do not need to change the construction of $\TT$.
		\item
		Derive the extended model existence and completeness theorems.
		\item
		How much classical logic did you use on the meta-level?
	\end{enumerate}
\end{exercise}

\begin{exercise}[Kripke Semantics]
	Prove the persistence and soundness for full Kripke semantics.
\end{exercise}

\begin{exercise}[Semantic Cut-Elimination]
	Consider the universal model $\UU$ over all contexts:
	\begin{enumerate}
		\item
		Relax the definition of Kripke forcing to accommodate inconsistent contexts as worlds of $\UU$.
		\item
		Establish semantic cut-elimination, i.e.~prove simultaneously:
		\begin{enumerate}
			\item
			$\UU\Vdash^\Gamma_\sigma A \text{ implies } \Gamma\vdash_{cf}A[\sigma]$
			\item
			$\text{If for all $\Delta\supseteq \Gamma$ and $B$ with $\Delta,A[\sigma]\vdash_{cf} B$ we have $\Delta\vdash_{cf} B$, then $\UU\Vdash^\Gamma_\sigma A$} $
		\end{enumerate}
	\end{enumerate}
\end{exercise}


\end{document}
